---
title: "Ach_Speed"
author: "Jiarui Fang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
Sys.setenv(LANGUAGE = "en")
```

```{r}
#library
library(tidyverse)
library(lme4)
library(lmerTest)
library(MuMIn)
library(performance)
library(readr)
library(dplyr)
library(ggeffects)
library(partR2)
library(slider)
library(tibble)
library(emmeans)
library(scales)
```

```{r load data}
# Vector of animal IDs (as character)
animal_ids <- c("GACH16", "GACH19", "GACH33", "GACH39", "GACH40")

# Base data directory
data_dir    <- "../data-raw"

# Empty list to collect all data.frames
data_list   <- list()

for (animal in animal_ids) {
  # build path to this animal’s top‐level folder
  animal_path <- file.path(data_dir, animal)
  
  # find all “*clean_data.csv” files anywhere under that folder
  csv_files   <- list.files(
    path        = animal_path,
    pattern     = "clean_data\\.csv$",
    recursive   = TRUE,
    full.names  = TRUE
  )
  
  # read each file and store in data_list
  for (file in csv_files) {
    df_name <- paste(animal, gsub(".*/", "", file), sep = "_")
    df_name # inspect the name of the data frame
    suppressMessages(df <- read_csv(file, show_col_types = FALSE))
    # remove any columns that has no header
    df <- df %>% select(-starts_with("..."))
    data_list[[df_name]] <- df
  }
}

```

```{r reformat data}
# Combine all data frames into one
all_data <- bind_rows(data_list)
# refomat the data
all_data <- all_data %>% 
  mutate(lightness = str_extract(session, "(?<=_)[^_]+$"))
all_data <- all_data %>% 
  mutate(session = paste(animal,lightness, date, sep = "_"))
all_data$session <- factor(all_data$session)
#remove rows containing NA in velocity or dFF_z
all_data <- all_data %>% drop_na(velocity, dFF_z)
all_data$lightness <- factor(all_data$lightness)
all_data$animal <- factor(all_data$animal)
rm(data_list, df)
all_data$lightness <- relevel(all_data$lightness, ref = "light")
all_data<- all_data %>% rename(speed = velocity) #it is actually velocity, wrong name in the original data
# create session_short variable for easier plotting
session_map <- all_data %>%
  distinct(animal, date, lightness, session) %>%  # unique session rows
  arrange(animal, date) %>%
  group_by(animal) %>%
  mutate(day_idx = dense_rank(date)) %>%          # same index for same date
  ungroup() %>%
  mutate(
    LD = if_else(lightness == "light", "L", "D"),
    session_facet = paste0(animal, "_", LD, "_", day_idx)
  )

all_data <- all_data %>%
  left_join(session_map %>% select(session, session_facet), by = "session")

all_data$session_facet <- factor(all_data$session_facet)
```

```{r binned data}
bin_width <- 0.2  # seconds; should match the GRAB ACh feature

binned_data <- all_data %>%
  arrange(animal, session, time) %>%
  mutate(
    bin = cut(
      time,
      breaks = seq(0, max(time, na.rm = TRUE) + bin_width, by = bin_width),
      right = FALSE,
      labels = FALSE
    ) - 1L
  ) %>%
  group_by(animal, session, lightness, bin) %>%
  summarise(
    # REMOVED time_bin calculation here
    dFF_z    = mean(dFF_z, na.rm = TRUE),
    speed = mean(speed, na.rm = TRUE),
    x        = mean(x, na.rm = TRUE),
    y        = mean(y, na.rm = TRUE),
    n_in_bin     = n(),
    session_facet = first(session_facet),
    .groups = "drop"
  ) %>%
  # ADD time_bin calculation here (it runs on the summarized rows)
  mutate(
    time_bin = (bin + 0.5) * bin_width
  ) %>%
  filter(n_in_bin >= 1)
# calculate change of speed (acceleration)
binned_data <- binned_data %>%
  group_by(session) %>%
  arrange(time_bin, .by_group = TRUE) %>%
  mutate(
    # central difference (better than forward difference)
    changeOfSpeed = (lead(speed) - lag(speed)) / (lead(time_bin) - lag(time_bin))
  ) %>%
  ungroup()
binned_data$log_speed <- log(binned_data$speed) # add log speed column
sum(!is.finite(binned_data$log_speed)) # check for log(0)=-Inf
#make session_facet ordered by LD, animal_id, idx
binned_data <- binned_data %>%
  mutate(
    LD = str_match(session_facet, "_([LD])_")[,2],                 # "L" or "D"
    animal_id = str_extract(session_facet, "^[^_]+"),              # "GACH16"
    idx = as.integer(str_match(session_facet, "_[LD]_(\\d+)$")[,2])# 1,2,3,...
  ) %>%
  arrange(LD, animal_id, idx) %>%
  mutate(session_facet = factor(session_facet, levels = unique(session_facet)))
```

```{r slow speed}
v_threshold <- 3   # slower than 3 cm/s
min_duration_s = 5

# binned_data has: session, time, speed (plus x, y, etc.)
bouts <- binned_data %>%
  arrange(session, time_bin) %>%
  group_by(session) %>%
  mutate(
    slow = speed < 3,
    bout_id = cumsum(slow != lag(slow, default = first(slow)))
  ) %>%
  group_by(session, bout_id) %>%
  summarise(
    slow = first(slow),
    t_start = first(time_bin),
    t_end   = last(time_bin),
    duration_s = as.numeric(t_end - t_start),
    .groups = "drop"
  ) %>%
  filter(slow, duration_s >= min_duration_s)

# Get all original rows that fall inside qualifying bouts
df_slow <- binned_data %>%
  arrange(session, time_bin) %>%
  group_by(session) %>%
  # Re-calculate bout_id on the full dataset
  mutate(
    slow = speed < 3,
    bout_id = cumsum(slow != lag(slow, default = first(slow)))
  ) %>%
  ungroup() %>%
  # ONLY keep rows where the (session, bout_id) pair exists in our filtered 'bouts' list
  semi_join(bouts, by = c("session", "bout_id"))

```
```{r immobility check}
max_distance_diameter <- 5  # cm

sampling_rate <- 1 / bin_width
window_size <- round(min_duration_s * sampling_rate)

df_immobile <- df_slow %>%
  arrange(session, time_bin) %>%
  group_by(session) %>%
  mutate(
    # 1) trailing window 内是否全部低速
    is_slow_window = slider::slide_lgl(
      speed, ~ all(.x < 3, na.rm = TRUE),
      .before = window_size - 1, .complete = TRUE
    ),

    # 2) trailing window 内的空间扩散是否 <= 5 cm（最大两点距离）
    is_small_area = slider::slide2_lgl(
      x, y,
      ~ {
        coords <- cbind(.x, .y)
        coords <- coords[stats::complete.cases(coords), , drop = FALSE]
        if (nrow(coords) < 2) return(FALSE)  # 没有足够点，不认为“immobile”
        max(stats::dist(coords)) <= max_distance_diameter
      },
      .before = window_size - 1, .complete = TRUE
    ),

    # 窗口末端点是否满足“低速+小范围”
    valid_end = is_slow_window & is_small_area
  ) %>%
  # 把“窗口末端点”为 TRUE 的窗口整体纳入（向前回看 window_size 个点）
  mutate(
    keep_row = slider::slide_lgl(
      valid_end, ~ any(.x),
      .before = window_size - 1, .complete = FALSE
    )
  ) %>%
  filter(keep_row) %>%
  select(-is_slow_window, -is_small_area, -valid_end, -keep_row)
```
```{r}
# 1. Create a variable for the folder path
output_folder <- "C:/Users/FangJ/OneDrive/研究 research/Research_Michael_Hasselmo/Ach/Paper/Figures"

# 2. Create the folder if it doesn't exist
if (!dir.exists(output_folder)) {
  dir.create(output_folder)
}
```
## 1. Visualization
```{r vis}
# Scatter plot of Velocity vs. dFF_z
set.seed(123)  # for reproducibility
p1 <- ggplot(
  binned_data %>% dplyr::sample_frac(0.05),
  aes(x = speed, y = dFF_z, color = lightness)
) +
  geom_point(size = 0.4, alpha = 0.6) +
  labs(
    x = "Speed (cm/s)",
    y = expression(Fluorescence  ~ (z-score))
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "speed_vs_dFF.tiff"), # Use .tiff or .pdf
  plot = p1,
  device = "tiff",        # Change to "pdf" if needed
  width = 180,            # Common width for 2-column journal images
  height = 120,           # Adjust height accordingly
  units = "mm",           # Use "mm" or "in"
  dpi = 300,              # Minimum requirement for most journals
  compression = "lzw",     # Reduces TIFF file size without losing quality
  bg = "white"            # Set background to white for better visibility
)
# Scatter plot of Speed vs. dFF_z faceted by sessions
p_speed_facet <- ggplot(binned_data, aes(x = speed, y = dFF_z, color = lightness)) +
  geom_point(size=0.5, alpha = 0.6) +
  facet_wrap(~ session_facet) +
  labs(x = "Speed (cm/s)", y = expression(Fluorescence  ~ (z-score))) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "speed_vs_dFF_faceted.tiff"),
  plot = p_speed_facet,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Scatter plot of Speed vs. dFF_z faceted by animals
p_speed_facet_animal <- ggplot(binned_data, aes(x = speed, y = dFF_z, color = lightness)) +
  geom_point(size=0.2, alpha = 0.2) +
  facet_wrap(~ animal) +
  labs(x = "Speed (cm/s)", y = expression(Fluorescence  ~ (z-score))) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "speed_vs_dFF_faceted_by_animal.tiff"),
  plot = p_speed_facet_animal,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
```{r vis log}
# Scatter plot of Log(speed) vs. dFF_z
p_logspeed <- ggplot(
  binned_data %>% dplyr::sample_frac(0.05),
  aes(x = log_speed, y = dFF_z, color = lightness)
) +
  geom_point(size = 0.4, alpha = 0.6) +
  labs(x = "Log(speed) (cm/s)",
    y = expression(Fluorescence  ~ (z-score))
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "log_speed_vs_dFF.tiff"),
  plot = p_logspeed,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

# Scatter plot of Log(speed) vs. dFF_z faceted by sessions
p_logspeed_facet <- ggplot(binned_data, aes(x = log_speed, y = dFF_z, color = lightness)) +
  geom_point(size=0.5, alpha = 0.6) +
  facet_wrap(~ session_facet) +
  labs(x = "Log(speed) (cm/s)", y = expression(Fluorescence  ~ (z-score))) +
  theme_minimal()

ggsave(
  filename = file.path(output_folder, "log_speed_vs_dFF_faceted.tiff"),
  plot = p_logspeed_facet,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

# Scatter plot of Log(speed) vs. dFF_z faceted by animals
p_logspeed_facet_animal <- ggplot(binned_data, aes(x = log_speed, y = dFF_z, color = lightness)) +
  geom_point(size=0.2, alpha = 0.2) +
  facet_wrap(~ animal) +
  labs(x = "Log(speed) (cm/s)", y = expression(Fluorescence  ~ (z-score))) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "log_speed_vs_dFF_faceted_by_animal.tiff"),
  plot = p_logspeed_facet_animal,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
## 2. Descriptive statistics
```{r}
# number of observations
n_obs <- nrow(binned_data)
cat("Number of observations:", n_obs, "\n")
n_imm_obs <- nrow(df_immobile)
cat("Number of immobility observations:", n_imm_obs, "\n")
cat("Percentage of immobility observations:", round((n_imm_obs / n_obs) * 100, 2), "%\n")

# A) lightness level
tbl_overall_by_lightness <- binned_data %>%
  group_by(lightness) %>%
  summarise(
    n_animals = n_distinct(animal),
    n_sessions = n_distinct(session),
    n_observations = n(),
    median_obs_per_session = median(as.numeric(table(session))),
    IQR_obs_per_session = IQR(as.numeric(table(session))),
    .groups = "drop"
  )

# B) animal level
tbl_sessions_per_animal <- binned_data %>%
  distinct(animal, session, lightness) %>%
  count(animal, lightness, name = "n_sessions") %>%
  pivot_wider(names_from = lightness, values_from = n_sessions, values_fill = 0) %>%
  mutate(
    has_both_conditions = if_else(rowSums(across(where(is.numeric)) > 0) >= 2, TRUE, FALSE)
  )

# C) session level
tbl_session_level <- binned_data %>%
  group_by(animal, session, lightness) %>%
  summarise(
    n_observations = n(),
    duration_s = max(time_bin, na.rm = TRUE) - min(time_bin, na.rm = TRUE),
    mean_speed = mean(speed, na.rm = TRUE),
    .groups = "drop"
  )

tbl_overall_by_lightness
tbl_sessions_per_animal
summary(tbl_session_level$n_observations)
summary(tbl_session_level$duration_s)

#comparison at session level
session_stats <- tbl_session_level %>%
  group_by(lightness) %>%
  summarise(
    mean_duration = mean(duration_s, na.rm = TRUE),
    sd_duration = sd(duration_s, na.rm = TRUE),
    mean_speed = mean(mean_speed, na.rm = TRUE),
    sd_speed = sd(mean_speed, na.rm = TRUE)
  )
# t-test for speed at session level between light and dark
t_test_session_speed <- t.test(mean_speed ~ lightness, data = tbl_session_level)
t_test_session_speed
```
```{r}
# Heatmaps of occupancy by sessions
p_occ_heatmap <- ggplot(binned_data, aes(x = x, y = y)) +
  geom_bin2d(aes(fill = after_stat(count / sum(count))), bins = 60) +
  scale_fill_viridis_c(name = "Occupancy (fraction)") +
  facet_wrap(~ session_facet) +
  coord_equal() +
  theme_minimal() +
  labs(x = "X position", y = "Y position")

ggsave(
  filename = file.path(output_folder, "occupancy_heatmap_by_session.tiff"),
  plot = p_occ_heatmap,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

# Heatmaps of speed by sessions
p_speed_heatmap <- ggplot(binned_data, aes(x = x, y = y, z = speed)) +
  stat_summary_2d(
    fun = mean,
    bins = 60
  ) +
  scale_fill_viridis_c(name = "Mean speed") +
  facet_wrap(~ session_facet) +
  coord_equal() +
  theme_minimal() +
  labs(
    x = "X position",
    y = "Y position"
  )
ggsave(
  filename = file.path(output_folder, "speed_heatmap_by_session.tiff"),
  plot = p_speed_heatmap,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Heatmaps of dFF_z by sessions
p_dFF_z_heatmap <- ggplot(binned_data, aes(x = x, y = y, z = dFF_z)) +
  stat_summary_2d(
    fun = mean,
    bins = 60
  ) +
  scale_fill_viridis_c(name = expression("Mean " * Fluorescence )) +
  facet_wrap(~ session_facet) +
  coord_equal() +
  theme_minimal() +
  labs(
    x = "X position",
    y = "Y position"
  )
ggsave(
  filename = file.path(output_folder, "dFF_z_heatmap_by_session.tiff"),
  plot = p_dFF_z_heatmap,
  device = "tiff",
  width = 250,
  height = 200,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Histogram of speed
p_hist_speed <- ggplot(binned_data, aes(x = speed,y = after_stat(count/sum(count)),fill = lightness)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ lightness) +
  labs(
       x = "Speed (cm/s)", 
       y = "Percentage") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "speed_distribution_by_lightness.tiff"),
  plot = p_hist_speed,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Histogram of dFF_z
p_hist_dFF_z <- ggplot(binned_data, aes(x = dFF_z, y = after_stat(count/sum(count)),fill = lightness)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~ lightness) +
  labs(
  x = expression(Fluorescence  ~ (z-score)), 
  y =  "Percentage") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "dFF_z_distribution_by_lightness.tiff"),
  plot = p_hist_dFF_z,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
```{r}
summary(binned_data$speed)
summary(binned_data$dFF_z)
t_test_speed <- t.test(speed ~ lightness, data = binned_data)
t_test_dFF_z <- t.test(dFF_z ~ lightness, data = binned_data)
cat("T-test for Speed:\n")
print(t_test_speed)
cat("T-test for dFF_z:\n")
print(t_test_dFF_z)
# Apply log transformation to speed
summary(binned_data$log_speed)
t_test_logspeed <- t.test(log_speed ~ lightness, data = binned_data)
cat("T-test for Log(speed):\n")
print(t_test_logspeed)
```
```{r}
# summary statistics by lightness
binned_data %>% 
  group_by(lightness) %>% 
  summarise(
    mean_dFF_z = mean(dFF_z),
    sd_dFF_z = sd(dFF_z),
    mean_speed = mean(speed),
    sd_speed = sd(speed),
    n = n()
  )

# one‐way ANOVA for velocity
res.aov_velocity <- aov(speed ~ animal, data = binned_data)
# overall F‐test
summary(res.aov_velocity)

# one‐way ANOVA for dFF_z
res.aov_dFF_z <- aov(dFF_z ~ animal, data = binned_data)
summary(res.aov_dFF_z)

# Statistics by animal
animal_stats <- binned_data %>%
  group_by(animal) %>%
  summarise(
    count = n(),
    mean_speed = mean(speed, na.rm = TRUE),
    sd_speed = sd(speed, na.rm = TRUE),
    mean_dFF_z = mean(dFF_z, na.rm = TRUE),
    sd_dFF_z = sd(dFF_z, na.rm = TRUE)
  )
print(animal_stats)
```


## 3. Regression based analyses

```{r linear}
# Linear mixed-effects model-Linear Relationship
model_lin <- lmer(dFF_z ~ speed + lightness + (1 | animal/session),
               data = binned_data, REML = FALSE)

# Summary of the model
summary(model_lin)

# 95% confidence intervals for fixed effects
confint(model_lin, parm = "beta_", method = "Wald")
```

```{r draw linear}
pred_lin <- ggpredict(model_lin, terms = c("speed [all]", "lightness"))
set.seed(123)
p_linear_sample<-ggplot(pred_lin, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Speed (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = binned_data %>% dplyr::sample_frac(0.05),
    aes(x = speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()

ggsave(
  filename = file.path(output_folder, "linear_model_dFF_vs_speed.tiff"),
  plot = p_linear_sample,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```

### Log transformation of speed
```{r log }
# Linear mixed-effects model with log-transformed speed
model_log <- lmer(dFF_z ~ log_speed + lightness + (1 | animal/session),
                  data = binned_data, REML = FALSE)

# Summarize the model
summary(model_log)

# 95% confidence intervals for fixed effects
confint(model_log, parm = "beta_", method = "Wald")

# LRT
anova(model_lin, model_log)

lme4::isSingular(model_log)
```


```{r draw log}
pred_log <- ggpredict(model_log, terms = c("log_speed [all]", "lightness"))
set.seed(123)
p_log_sample <- ggplot(pred_log, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Log(speed) (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = binned_data %>% dplyr::sample_frac(0.05),
    aes(x = log_speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "log_model_dFF_vs_log_speed.tiff"),
  plot = p_log_sample,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```


## 4. Interaction model
```{r log only & middle random effect}
model_log_interaction_mid <- lmer(dFF_z ~ log_speed * lightness + (1 | animal:session) + (log_speed || animal),
           data = binned_data, REML = FALSE)
isSingular(model_log_interaction_mid) # if any random effect is singular
VarCorr(model_log_interaction_mid)
anova(model_log, model_log_interaction_mid)
model_log_interaction_mid_REML<- update(model_log_interaction_mid,
                                             REML = TRUE)
summary(model_log_interaction_mid_REML)
confint(model_log_interaction_mid_REML, parm = "beta_", method = "Wald")

# A) slopes for each lightness level
slopes <- emtrends(model_log_interaction_mid_REML, ~ lightness, var = "log_speed")
slopes
confint(slopes)

# B) pairwise comparisons of slopes
slope_contrast <- pairs(slopes)
slope_contrast
confint(slope_contrast)

slopes_tbl <- summary(slopes) %>% as.data.frame()
slope_contrast_tbl <- summary(slope_contrast) %>% as.data.frame()
slopes_tbl
slope_contrast_tbl
```
Good, this is not singular, and the interaction is still significant.

```{r draw log interaction mid}
pred_log_interaction_final <- ggpredict(model_log_interaction_mid_REML, terms = c("log_speed [all]", "lightness"))
set.seed(123)
p_log_interaction <- ggplot(pred_log_interaction_final, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Log(speed) (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = binned_data %>% dplyr::sample_frac(0.05),
    aes(x = log_speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "log_interaction_model_dFF_vs_log_speed.tiff"),
  plot = p_log_interaction,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

pred_raw <- as.data.frame(pred_log_interaction_final) %>%
  mutate(
    x = as.numeric(x),
    speed = exp(x)
  )

# use the raw speed for x-axis but plot the log-based model predictions
set.seed(123)
p_log_interaction_rawx <- ggplot(pred_raw, aes(x = speed, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_point(
    data = binned_data %>% dplyr::sample_frac(0.05),
    aes(x = speed, y = dFF_z, color = lightness),
    size = 0.2, alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    x = "Speed (cm/s)",
    y = expression(Fluorescence~(z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "log_interaction_model_dFF_vs_speed_rawx.tiff"),
  plot = p_log_interaction_rawx,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

```
```{r linear only & middle random effect}
model_lin_interaction_mid <- lmer(dFF_z ~ speed * lightness + (1 | animal:session) + (speed || animal),
           data = binned_data, REML = FALSE)
isSingular(model_lin_interaction_mid) # if any random effect is singular
VarCorr(model_lin_interaction_mid)
anova(model_lin, model_lin_interaction_mid)
model_lin_interaction_mid_REML<- update(model_lin_interaction_mid,
                                             REML = TRUE)
summary(model_lin_interaction_mid_REML)
confint(model_lin_interaction_mid_REML, parm = "beta_", method = "Wald")
```
```{r draw linear interaction mid}
pred_lin_interaction_final <- ggpredict(model_lin_interaction_mid_REML, terms = c("speed [all]", "lightness"))
set.seed(123)
p_lin_interaction <- ggplot(pred_lin_interaction_final, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Speed (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = binned_data %>% dplyr::sample_frac(0.05),
    aes(x = speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "linear_interaction_model_dFF_vs_speed.tiff"),
  plot = p_lin_interaction,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
```{r  LOSO for log vs linear}
dat_cv <- binned_data %>%
  mutate(
    lightness    = droplevels(lightness),
    animal       = droplevels(animal),
    session      = droplevels(session)
  ) %>%
  filter(
    !is.na(dFF_z),
    !is.na(lightness),
    !is.na(animal),
    !is.na(session)
  )

sessions <- levels(dat_cv$session)

score_fold <- function(model, test_df, y = "dFF_z", pred_fixed_only = TRUE) {
  # pred_fixed_only=TRUE：only use fixed effects for prediction
  # pred_fixed_only=FALSE：use conditional prediction (including random effects of animal; new levels of animal:session will be set to 0)
  pred <- predict(
    model,
    newdata = test_df,
    re.form = if (pred_fixed_only) NA else NULL,
    allow.new.levels = TRUE
  )
  err  <- test_df[[y]] - pred
  rmse <- sqrt(mean(err^2, na.rm = TRUE))
  mae  <- mean(abs(err), na.rm = TRUE)

  # Gaussian log score (the bigger the better)
  sig <- sigma(model)
  logscore <- mean(dnorm(test_df[[y]], mean = pred, sd = sig, log = TRUE), na.rm = TRUE)

  list(rmse = rmse, mae = mae, logscore = logscore)
}

# 1) LOSO：leave-one-session-out cross-validation
results_loso <- lapply(sessions, function(s_out) {
  train_df <- dat_cv %>% filter(session != s_out)
  test_df  <- dat_cv %>% filter(session == s_out)

  # because we want to compare linear vs. log models fixed effects, let REML=FALSE
  m_lin <- lmer(
    dFF_z ~ speed * lightness + (1 | animal:session) + (speed || animal),
    data = train_df, REML = FALSE
  )

  m_log <- lmer(
    dFF_z ~ log_speed * lightness + (1 | animal:session) + (log_speed || animal),
    data = train_df, REML = FALSE
  )

  sc_lin <- score_fold(m_lin, test_df, pred_fixed_only = TRUE)
  sc_log <- score_fold(m_log, test_df, pred_fixed_only = TRUE)

  tibble(
    session = s_out,
    n_test  = nrow(test_df),

    rmse_lin = sc_lin$rmse,
    rmse_log = sc_log$rmse,
    d_rmse   = sc_log$rmse - sc_lin$rmse,     # <0 favors log

    mae_lin  = sc_lin$mae,
    mae_log  = sc_log$mae,
    d_mae    = sc_log$mae - sc_lin$mae,       # <0 favors log

    lscore_lin = sc_lin$logscore,
    lscore_log = sc_log$logscore,
    d_lscore   = sc_log$logscore - sc_lin$logscore  # >0 favors log
  )
}) %>% bind_rows()

# 2) summary of LOSO results
summary_loso <- results_loso %>%
  summarise(
    n_sessions = n(),
    mean_d_rmse = mean(d_rmse),
    sd_d_rmse   = sd(d_rmse),
    mean_d_mae  = mean(d_mae),
    sd_d_mae    = sd(d_mae),
    mean_d_lscore = mean(d_lscore),
    sd_d_lscore   = sd(d_lscore)
  )

print(summary_loso)

# 3) t.test is actually wrong, but for reference
# d_rmse < 0 --> log and d_lscore > 0 --> log
t_rmse   <- t.test(results_loso$d_rmse, mu = 0)
t_mae    <- t.test(results_loso$d_mae,  mu = 0)
t_lscore <- t.test(results_loso$d_lscore, mu = 0)

t_rmse
t_mae
t_lscore

# 4) bootstrap 95% CI for mean differences
set.seed(1)
B <- 5000
boot_mean_drmse <- replicate(B, mean(sample(results_loso$d_rmse, replace = TRUE)))
boot_ci_drmse <- quantile(boot_mean_drmse, c(0.025, 0.975))

boot_mean_dlscore <- replicate(B, mean(sample(results_loso$d_lscore, replace = TRUE)))
boot_ci_dlscore <- quantile(boot_mean_dlscore, c(0.025, 0.975))

boot_ci_drmse
boot_ci_dlscore
```
```{r effect size}
# effect size
r.squaredGLMM(model_lin)
r.squaredGLMM(model_log)
r.squaredGLMM(model_lin_interaction_mid_REML)
r.squaredGLMM(model_log_interaction_mid_REML)
AIC(model_lin, model_log)
AIC(model_lin_interaction_mid, model_log_interaction_mid)
```

```{r check change of speed}
binned_data_clear <- binned_data %>%
  drop_na(changeOfSpeed)
model_log_only <- lmer(dFF_z ~ log_speed + lightness + (1 | animal:session) + (log_speed || animal),
           data = binned_data_clear, REML = FALSE, na.action = na.omit)
model_ChangeOfSpeedAndLog <- lmer(dFF_z ~ log_speed * lightness + changeOfSpeed * lightness + (1 | animal:session) + (log_speed || animal) + (changeOfSpeed || animal),
           data = binned_data_clear, REML = FALSE, na.action = na.omit)
summary(model_ChangeOfSpeedAndLog)
confint(model_ChangeOfSpeedAndLog, parm = "beta_", method = "Wald")
anova(model_log_only, model_ChangeOfSpeedAndLog)

model_absChangeOfSpeedAndLog <- lmer(dFF_z ~ log_speed * lightness + abs(changeOfSpeed) * lightness + (1 | animal:session) + (log_speed || animal) + (abs(changeOfSpeed) || animal),
           data = binned_data_clear, REML = FALSE, na.action = na.omit)
summary(model_absChangeOfSpeedAndLog)
confint(model_absChangeOfSpeedAndLog, parm = "beta_", method = "Wald")
anova(model_log_only, model_absChangeOfSpeedAndLog)
check_collinearity(model_ChangeOfSpeedAndLog)
check_collinearity(model_absChangeOfSpeedAndLog)
```
```{r effect size2}
# effect size
model_ChangeOfSpeedAndLog_REML <- update(model_ChangeOfSpeedAndLog, REML = TRUE)
r.squaredGLMM(model_ChangeOfSpeedAndLog_REML)
model_absChangeOfSpeedAndLog_REML <- update(model_absChangeOfSpeedAndLog, REML = TRUE)
r.squaredGLMM(model_absChangeOfSpeedAndLog_REML)
model_log_only_REML <- update(model_log_only, REML = TRUE)
r.squaredGLMM(model_log_only_REML)
AIC(model_log_only, model_ChangeOfSpeedAndLog, model_absChangeOfSpeedAndLog)
```


## 5. regression with immobility
```{r immobility model}
imm_linear_interaction_mid <- lmer(dFF_z ~ speed * lightness + (1 | animal:session) + (speed || animal),
               data = df_immobile, REML = FALSE)
imm_log_interaction_mid <- lmer(dFF_z ~ log_speed * lightness + (1 | animal:session) + (log_speed || animal),
               data = df_immobile, REML = FALSE)
```
```{r}
summary(imm_linear_interaction_mid)
summary(imm_log_interaction_mid)
anova(imm_linear_interaction_mid, imm_log_interaction_mid)
isSingular(imm_log_interaction_mid)
confint(imm_log_interaction_mid, parm = "beta_", method = "Wald")
confint(imm_linear_interaction_mid, parm = "beta_", method = "Wald")
```
```{r}
imm_log_interaction_mid_REML <- update(imm_log_interaction_mid, REML = TRUE)
summary(imm_log_interaction_mid_REML)
imm_linear_interaction_mid_REML <- update(imm_linear_interaction_mid, REML = TRUE)
summary(imm_linear_interaction_mid_REML)
```
```{r effect size imm}
# effect size
r.squaredGLMM(imm_linear_interaction_mid_REML)
r.squaredGLMM(imm_log_interaction_mid_REML)
AIC(imm_linear_interaction_mid, imm_log_interaction_mid)
```
```{r draw linear imm}
pred_lin <- ggpredict(imm_linear_interaction_mid_REML, terms = c("speed [all]", "lightness"))
set.seed(123)
p_imm_linear_sample<-ggplot(pred_lin, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Speed (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = df_immobile %>% dplyr::sample_frac(0.05),
    aes(x = speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()

ggsave(
  filename = file.path(output_folder, "imm_linear_interaction_model_dFF_vs_speed.tiff"),
  plot = p_imm_linear_sample,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
```{r draw log interaction imm}
pred_log_interaction_final <- ggpredict(imm_log_interaction_mid_REML, terms = c("log_speed [all]", "lightness"))
set.seed(123)
p_imm_log_interaction <- ggplot(pred_log_interaction_final, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    x = "Log(speed) (cm/s)",
    y = expression(Fluorescence  ~ (z-score)),
    color = "Lightness",
    fill  = "Lightness"
  ) +
  geom_point(
    data = df_immobile %>% dplyr::sample_frac(0.2),
    aes(x = log_speed, y = dFF_z, color = lightness),
    size = 0.2,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "imm_log_interaction_model_dFF_vs_log_speed.tiff"),
  plot = p_imm_log_interaction,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
```
```{r  imm LOSO for log vs linear}
dat_cv <- df_immobile %>%
  mutate(
    lightness    = droplevels(lightness),
    animal       = droplevels(animal),
    session      = droplevels(session)
  ) %>%
  filter(
    !is.na(dFF_z),
    !is.na(lightness),
    !is.na(animal),
    !is.na(session)
  )

sessions <- levels(dat_cv$session)

score_fold <- function(model, test_df, y = "dFF_z", pred_fixed_only = TRUE) {
  # pred_fixed_only=TRUE：only use fixed effects for prediction
  # pred_fixed_only=FALSE：use conditional prediction (including random effects of animal; new levels of animal:session will be set to 0)
  pred <- predict(
    model,
    newdata = test_df,
    re.form = if (pred_fixed_only) NA else NULL,
    allow.new.levels = TRUE
  )
  err  <- test_df[[y]] - pred
  rmse <- sqrt(mean(err^2, na.rm = TRUE))
  mae  <- mean(abs(err), na.rm = TRUE)

  # Gaussian log score (the bigger the better)
  sig <- sigma(model)
  logscore <- mean(dnorm(test_df[[y]], mean = pred, sd = sig, log = TRUE), na.rm = TRUE)

  list(rmse = rmse, mae = mae, logscore = logscore)
}

# 1) LOSO：leave-one-session-out cross-validation
results_loso <- lapply(sessions, function(s_out) {
  train_df <- dat_cv %>% filter(session != s_out)
  test_df  <- dat_cv %>% filter(session == s_out)

  # because we want to compare linear vs. log models fixed effects, let REML=FALSE
  m_lin <- lmer(
    dFF_z ~ speed * lightness + (1 | animal:session) + (speed || animal),
    data = train_df, REML = FALSE
  )

  m_log <- lmer(
    dFF_z ~ log_speed * lightness + (1 | animal:session) + (log_speed || animal),
    data = train_df, REML = FALSE
  )

  sc_lin <- score_fold(m_lin, test_df, pred_fixed_only = TRUE)
  sc_log <- score_fold(m_log, test_df, pred_fixed_only = TRUE)

  tibble(
    session = s_out,
    n_test  = nrow(test_df),

    rmse_lin = sc_lin$rmse,
    rmse_log = sc_log$rmse,
    d_rmse   = sc_log$rmse - sc_lin$rmse,     # <0 favors log

    mae_lin  = sc_lin$mae,
    mae_log  = sc_log$mae,
    d_mae    = sc_log$mae - sc_lin$mae,       # <0 favors log

    lscore_lin = sc_lin$logscore,
    lscore_log = sc_log$logscore,
    d_lscore   = sc_log$logscore - sc_lin$logscore  # >0 favors log
  )
}) %>% bind_rows()

# 2) summary of LOSO results
summary_loso <- results_loso %>%
  summarise(
    n_sessions = n(),
    mean_d_rmse = mean(d_rmse),
    sd_d_rmse   = sd(d_rmse),
    mean_d_mae  = mean(d_mae),
    sd_d_mae    = sd(d_mae),
    mean_d_lscore = mean(d_lscore),
    sd_d_lscore   = sd(d_lscore)
  )

print(summary_loso)

# 3) t.test is actually wrong, but for reference
# d_rmse < 0 --> log and d_lscore > 0 --> log
t_rmse   <- t.test(results_loso$d_rmse, mu = 0)
t_mae    <- t.test(results_loso$d_mae,  mu = 0)
t_lscore <- t.test(results_loso$d_lscore, mu = 0)

t_rmse
t_mae
t_lscore

# 4) bootstrap 95% CI for mean differences
set.seed(1)
B <- 5000
boot_mean_drmse <- replicate(B, mean(sample(results_loso$d_rmse, replace = TRUE)))
boot_ci_drmse <- quantile(boot_mean_drmse, c(0.025, 0.975))

boot_mean_dlscore <- replicate(B, mean(sample(results_loso$d_lscore, replace = TRUE)))
boot_ci_dlscore <- quantile(boot_mean_dlscore, c(0.025, 0.975))

boot_ci_drmse
boot_ci_dlscore
```
## 6. Methods for examining regression assumptions

```{r}
binned_data$fitted_values <- fitted(model_log_interaction_mid_REML)
binned_data$residuals <- residuals(model_log_interaction_mid_REML)

# Scatter plot of residuals vs. fitted values
p_residuals_vs_fitted <- ggplot(binned_data, aes(x = fitted_values, y = residuals, color = lightness)) +
  geom_point(size=0.2, alpha = 0.2) +
  labs(x = "Fitted Fluorescence (z-score)", y = "Residuals") +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "residuals_vs_fitted.tiff"),
  plot = p_residuals_vs_fitted,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Scatter plot of residuals vs. log_speed
p_residuals_vs_log_speed <- ggplot(binned_data, aes(x = log_speed, y = residuals, color = lightness)) +
  geom_point(size=0.2, alpha = 0.2) +
  geom_smooth(se = FALSE) +
  labs(x = "Log(speed) (cm/s)", y = "Residuals") +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "residuals_vs_log_speed.tiff"),
  plot = p_residuals_vs_log_speed,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Scatter plot of observed vs. fitted values
p_observed_vs_fitted <- ggplot(binned_data, aes(x = fitted_values, y = dFF_z, color = lightness)) +
  geom_point(size=0.2, alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "Fitted Fluorescence (z-score)", 
       y = "Observed Fluorescence (z-score)") +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "observed_vs_fitted.tiff"),
  plot = p_observed_vs_fitted,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)
# Q-Q plot of residuals
p_qq_plot <- ggplot(binned_data, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
ggsave(
  filename = file.path(output_folder, "qq_plot_residuals.tiff"),
  plot = p_qq_plot,
  device = "tiff",
  width = 180,
  height = 150,
  units = "mm",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

```


```{r autocorrelation}
# compute lag-1 autocorrelation per session, then summarize
acf_by_session <- binned_data %>%
  group_by(animal, session) %>%
  arrange(time_bin, .by_group = TRUE) %>%
  summarise(
    acf1 = cor(residuals[-1], residuals[-n()], use = "complete.obs"),
    .groups = "drop"
  )

summary(acf_by_session$acf1)
```
It is highly autocorrelated, which is not surprising but not ideal for LMM. But given the complexity of the data structure, it is hard to model the autocorrelation properly, and other researchers have also used LMM for similar data.